---
title: "Target trial emulation"
author: "maj"
date: "2025-05-11"
date-modified: last-modified
bibliography: ../etc/refs.bib
csl: ../etc/biomed-central.csl
categories:
  - causal-inference
  - bayesian
---

https://www.youtube.com/watch?v=vtlxc4MG4zs

https://www.youtube.com/watch?v=9lMhD2VfEjI


Three main sources of bias

Confounding
Immortal time
Selection


# Introduction

What's the causal question?

What is the target trial protocol in terms of:

+ eligibility, 
+ treatment options (including start and end times), 
+ assignment procedures
+ follow up
+ outcome of interest
+ causal contrasts and estimation procedure


+ Specify the target trial — eligibility, interventions, follow-up, outcome, estimand
+ Define our observational analog — map each trial element to registry data
+ Construct analytic dataset — clone, censor, weight (IPTW)
+ Fit a Bayesian outcome model with cmdstanr
+ Interpret the estimated causal effect of antibiotic duration on cure at day 7

Assume the following trial setup:

| Trial Element     | This Registry Trial Emulation                            |
| :---------------- | :--------------------------------------------------------|
| **Eligibility**   | Children aged 6 m–5 y with acute respiratory infection   |
| **Time zero**     | Day of clinical assessment and antibiotic decision       |
| **Interventions** | Days of antibiotics prescribed                           |
| **Follow-up**     | 7 days post–time zero                                    |
| **Outcome**       | Cure by day 7 (yes/no)                                   |
| **Estimand**      | Risk difference ITT by successive days of therapy        |

Above is speculative. How do we actually conceptualise intervention?


To emulate randomisation, need to account for baseline and time-varying confounders.


Day treatment commenced (prescription filled and medication taken per directions).

Assume the registry has the following information on each individual that was enrolled.

+ id
+ age_mo
+ sex
+ viral_test (positive/negative)
+ bacterial_test (positive/negative)
+ duration_symptoms_prior (days before visit)
+ abx_dur (observed days of antibiotic prescription; 0 to 7)
+ cured7 (1 = cured by day 7, 0 = not)

plus baseline covariates such as comorbidities, maybe a severity score, prior antibiotic exposure etc

To keep things simple, collapse abx_dur into three durations: 

+ 0 days (none)
+ 3 days (1–2 or 3–6 months grouped)
+ 7 days (7+)

Simulate the registry data (this is suppose to represent what we actually collected).
Again, keep it relatively simple and ignore the possibility of time-varying data, censoring and non-adherence.

```{r}
library(data.table)
library(ggplot2)

# Sample size
N <- 3129

# over represent lower age class
xx <- exp(rnorm(55))
p_age_mo <- xx / sum(xx)
p_age_mo <- sort(p_age_mo, decreasing = T)

# Simulate baseline variables
d_orig <- data.table(
  id = 1:N,
  # in practice the db would contain a broader age range
  age_mo = sample(6:60, N, replace = TRUE, prob = p_age_mo),
  sex = rbinom(N, 1, 0.5),  # 0 = female, 1 = male
  viral_test = rbinom(N, 1, 0.4),      # +ve viral test
  bacterial_test = rbinom(N, 1, 0.35), # +ve bact test
  duration_symptoms_prior = pmin(7, rpois(N, 2)), # capped at 7
  severity_score = round(rnorm(N, mean = 1, sd = 1), 0)
)

# abx prescribed
d_orig[, linpred_abx := -1 +
         1.4 * bacterial_test +
         2.2 * severity_score +
         0.4 * scale(age_mo) +
         0.2 * duration_symptoms_prior +
         0.3 * (1 - viral_test) ]

# Convert to probability for any antibiotic use
d_orig[, p_abx := plogis(linpred_abx)]

# Assign antibiotic prescribed or not
d_orig[, abx_any := rbinom(.N, 1, p_abx)]

# dcast(d_orig[, .N, keyby = .(abx_any, bacterial_test)], bacterial_test ~ abx_any, value.var = "N")
# dcast(d_orig[, .N, keyby = .(abx_any, severity_score)], severity_score ~ abx_any, value.var = "N")

# Abx duration (in days) conditional on abx_any
xx <- 1 - 2 *exp(-c(1:7))
p_abx_dur <- xx / sum(xx)
p_abx_dur <- sort(p_abx_dur, decreasing = F)

d_orig[, abx_dur := 0L]
d_orig[abx_any == 1, abx_dur := sample(
  1:7, .N, replace = TRUE, prob = p_abx_dur
)]

# d_orig[, .N, keyby = abx_any]
# d_orig[, .N, keyby = abx_dur]

f_emax <- function(
    x = 0:7, 
    e0 = qlogis(0.2), 
    emax = qlogis(0.8) - qlogis(0.2), 
    ec50 = 2, 
    h = 1.15, 
    prob = T){
  
  eta <- e0 +  (emax * x^h)/ (ec50^h + x^h)
  if(prob){
    return(plogis(eta))
  } else{
    return(eta)
  }
}

d_orig[
  bacterial_test == T, 
  linpred_cure := 
    f_emax(abx_dur, prob = F) - 0.1 * severity_score - 0.1 * (viral_test == 1)]

# ggplot(d_orig[bacterial_test == T], aes(x = abx_dur, y = plogis(linpred_cure))) +
#   geom_point()

d_orig[
  bacterial_test == F, 
  linpred_cure := rnorm(.N, -0.4 - 0.1 * severity_score - 0.1 * (viral_test == 1), 0.1)]

# ggplot(d_orig[bacterial_test == F], aes(x = abx_dur, y = plogis(linpred_cure))) +
#   geom_point()

# Convert to probability and simulate outcome
d_orig[, p_cure := plogis(linpred_cure)]
d_orig[, cured7 := rbinom(.N, 1, p_cure)]

# Clean up
d_orig[, c("linpred_abx", "p_abx", "linpred_cure", "p_cure", "abx_any") := NULL]

# View structure
d_orig[]

```









```{r}
library(data.table)

d_1 <- copy(d_orig)

#— Define treatment strategies
d_1[, trt := fifelse(abx_dur == 0, 0L,
             fifelse(abx_dur <= 6, 3L, 7L))]

#— Baseline covariates
Wvars <- c("age_mo", "sex", "viral_test", "bacterial_test",
           "duration_symptoms_prior", "severity_score")

#— Fit multinomial propensity model
library(nnet)
ps_mod <- multinom(trt ~ ., data = d_1[, c("trt", Wvars), with=FALSE])

#— Predict propensity
p_mat <- predict(ps_mod, type = "probs")
d_1[, paste0("ps0_", c(0,3,7)) := as.data.table(p_mat)]
# weights: 1 / P(A=observed | W)
d_1[, ipw := 1 / fifelse(trt==0, ps0_0,
                 fifelse(trt==3, ps0_3, ps0_7))]

#— Stabilized weights (optional)
p_dist <- prop.table(table(d_1$trt))
d_1[, sw := fifelse(trt==0, p_dist["0"]/ps0_0,
            fifelse(trt==3, p_dist["3"]/ps0_3, p_dist["7"]/ps0_7))]

#— Inspect weights
d_1[, .(mean_ipw = mean(ipw), mean_sw = mean(sw))]

```

