---
title: "Stan implementation of AFT survival model with log-logistic event times"
author: "maj"
date: "2025-03-13"
date-modified: last-modified
bibliography: ../etc/refs.bib
csl: ../etc/biomed-central.csl
embed-resources: true
categories:
  - survival analysis
  - bayesian
---

```{r}
#| code-fold: true

library(data.table)
library(ggplot2)
library(gt)
```

The hazard function associated with log-logistic event times is hump-shaped, a bit like theand is a bit like a log normal but with longer tails.
It initially increases, reaches a maximum and then decreases toward 0 as lifetimes become larger and larger.
Definitions for the density function can be found in the stan docs:
[https://mc-stan.org/docs/functions-reference/positive_continuous_distributions.html#log-logistic-distribution](https://mc-stan.org/docs/functions-reference/positive_continuous_distributions.html#log-logistic-distribution) and in the `flexsurv` help file, see `?flexsurv::dllogis`.
Unlike lognormal, loglistic has a closed form hazard function.

$$
\begin{aligned}
f = \frac{(\beta/\alpha)(t/\alpha)^{\beta-1}}{(1 + (t/\alpha)^\beta)^2}
\end{aligned}
$$

the cumulative distribution function is

$$
\begin{aligned}
F = \frac{1}{1 + (t/\alpha)^{-\beta}}
\end{aligned}
$$

the survival function is $1 - F$:

$$
\begin{aligned}
S &= 1 - \frac{1}{1 + (t/\alpha)^{-\beta}} \\
  &= \frac{1 + (t/\alpha)^{-\beta}}{1 + (t/\alpha)^{-\beta}} - \frac{1}{1 + (t/\alpha)^{-\beta}} \\
  &= \frac{(t/\alpha)^{-\beta}}{1 + (t/\alpha)^{-\beta}} \\
  &= \frac{1}{(t/\alpha)^\beta (1 + (t/\alpha)^{-\beta})} \\
  &= \frac{1}{1 + (t/\alpha)^{\beta}} \\
\end{aligned}
$$

the hazard function is $f/S$:

$$
\begin{aligned}
h &= \frac{\frac{(\beta/\alpha)(t/\alpha)^{\beta-1}}{(1 + (t/\alpha)^\beta)^2}}{\frac{1}{1 + (t/\alpha)^{\beta}}} \\
  &= \frac{(\beta/\alpha)(t/\alpha)^{\beta-1} (1 + (t/\alpha)^{\beta}) }{(1 + (t/\alpha)^\beta)^2} \\
  &= \frac{(\beta/\alpha)(t/\alpha)^{\beta-1}}{1 + (t/\alpha)^\beta} \quad \text{cancelling similar terms}
\end{aligned}
$$

Say that we want to simulate data where there was 10% cumulative incidence by day 360, e.g. in the first 360 days of life about 10% of infants will experience a medical attendance for RSV, a respiratory illness.

We want $S(360) = \frac{1}{1 + (360/\alpha)^\beta} = \pi = 0.9$.
As $\alpha$ is the scale parameter, which is modelled as a linear function of parameters (treatment effects etc), assume that $\beta$ is known and solve for $\alpha$

$$
\begin{aligned}
\alpha = \frac{360}{((1/\pi) - 1)^{1/\beta}}
\end{aligned}
$$

For example, say $\beta = 2$, this implies $\alpha = \frac{360}{((1/0.9) - 1)^{1/2}} = 1080$ gives

```{r}
#| code-fold: true
#| layout-ncol: 2
#| layout-nrow: 2
#| lightbox: true

# log-logistic parameters
# shape parameter
b <- 2
# scale
a <- 360 / ( (1/0.9)-1 )^(1/b)    

# Create a data.table with days from 1 to 360
dt <- data.table(day = 1:360)

# Compute the survival function S(t) = 1 / (1 + (t/a)^b)
dt[, survival := 1 / (1 + (day / a)^b)]

# Compute the density f(t) = (gamma/alpha) * (t/alpha)^(gamma-1) / [1 + (t/alpha)^gamma]^2
dt[, density := (b / a) * (day / a)^(b - 1) / (1 + (day / a)^b)^2]

# Compute the hazard function h(t) = f(t) / S(t)
dt[, hazard := density / survival]

# Plot the survival curve
ggplot(dt, aes(x = day, y = survival)) +
  geom_line(color = "blue", lwd = 0.4) +
  labs(title = "Survival Curve for RSV (Log-Logistic Model)",
       x = "Day of Life", y = "Survival Probability S(t)") +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal()

# Plot the hazard function
ggplot(dt, aes(x = day, y = hazard)) +
  geom_line(color = "darkgreen", lwd = 0.4) +
  labs(title = "Hazard Function for RSV (Log-Logistic Model)",
       x = "Day of Life", y = "Hazard h(t)") +
  theme_minimal()

# Plot the density function
ggplot(dt, aes(x = day, y = density)) +
  geom_line(color = "red", lwd = 0.4) +
  labs(title = "Density Function for RSV (Log-Logistic Model)",
       x = "Day of Life", y = "Density f(t)") +
  theme_minimal()
```

In AFT, the scale parameter is allowed to vary with the covariates, such as:

$$
\begin{aligned}
\alpha_i &= \exp(\mu_i) \\
\mu &= \gamma_0 + \gamma_1 x_1 + \dots
\end{aligned}
$$

The density for observation $i$ is then:

$$
\begin{aligned}
f(t_i) &= \frac{\beta}{\exp(\mu_i)} \left( \frac{t_i}{\exp(\mu_i)}\right)^{\beta - 1} \left[  1 + \left(\frac{t_i}{\exp(\mu_i)} \right)^\beta \right]^{-2}
\end{aligned}
$$

taking logs of this gives the log-likelihood for observation $i$:

$$
\begin{aligned}
\log f(t_i) &= \log(\beta) - \mu_i + (\beta - 1)\left[ \log(t_i) - \mu_i \right]   -2 \left( 1 + \exp(  \beta [\log(t_i) - \mu_i] )   \right)
\end{aligned}
$$

for the censored observations, the survival function is used:

$$
\begin{aligned}
S = \frac{1}{1 + (t_i/\exp(\mu_i))^\beta}
\end{aligned}
$$

taking logs:

$$
\begin{aligned}
\log S &= 0 + \log \left[ 1 + \left( \frac{t_i}{\exp(\mu_i)} \right)^\beta        \right] \\
  &= \log \left[ 1 + \exp ( \beta (\log (t_i) - \mu_i)    )    \right] \\ 
\end{aligned}
$$

Implement stan model:

```{r}
#| class-output: stan
#| echo: false

cat(readLines("stan/log-logistic-aft-01.stan"), sep = "\n")
```

Running the model with the assumed data gives parameter estimates.

```{r, eval = T}
#| code-fold: true

mod_01 <- cmdstanr::cmdstan_model("stan/log-logistic-aft-01.stan")

# Simulation parameters
N <- 500  # Number of observations
P <- 2    # Number of predictors

gamma_true <- c(log(1080), 1)  
# True shape parameter
shape_true <- 2  

# Simulate covariates
simulate_data <- function() {
  
  d <- data.table(
    x1 = rep(0:1, length  = N)
  )
  
  # Design matrix
  X <- as.matrix(d[, .(1, x1)])
  
  # Linear predictor
  mu <- X %*% gamma_true
  
  # Scale parameter (a in the formula)
  scale <- exp(mu)
  
  # Simulate survival times from log-logistic distribution
  # Using quantile function: t = a * (p/(1-p))^(1/b) where p is uniform(0,1)
  u <- runif(N)
  t <- scale * (u / (1 - u))^(1 / shape_true)
  
  # Simulate censoring (administrative censoring at a fixed time point)
  max_follow_up <- quantile(t, 0.8)  # Censoring at 80th percentile of times
  censored <- t > max_follow_up
  observed_time <- pmin(t, max_follow_up)
  event <- as.numeric(!censored)
  
  # Return data
  d[, `:=`(
    time = observed_time,
    event = event
  )]
  
  return(list(
    data = d,
    X = X,
    true_times = t,
    censoring_rate = mean(censored)
  ))
}

# Simulate data
sim_data <- simulate_data()


cat("Censoring rate:", sim_data$censoring_rate, "\n")

# Prepare data for Stan
ld <- list(
  N = N,
  P = P,
  X = sim_data$X,
  y = sim_data$data$time,
  event = sim_data$data$event,
  N_pred = 361,
  t_surv = 0:360
)


# Fit the Stan model - sink to remove the noise
snk <- capture.output(
  m1 <- mod_01$sample(
      ld, iter_warmup = 1000, iter_sampling = 1000,
      parallel_chains = 4, chains = 4, refresh = 0, show_exceptions = F,
      max_treedepth = 10)
)

```

Extract the parameters that we are interested in:

```{r}
#| code-fold: true

m1$summary(variables = c("gamma", "shape"))
```

Exponentiating the $\gamma_2$ parameter gives the acceleration factor associated with the treatment effect.
For example, if $\gamma_2 > 0$ we can say that change from the control to treatment arm is associated with survival times being multiplied by a factor of $\exp(\gamma_2)$, indicating prolonged survival.
Similarly, if $\gamma_2 < 0$ we have a reduction in survival (the time to event speeds up).

In a log-logistic AFT model, I think that the median survival time for an individual with covariates $X$ is given by $\exp( X' \gamma)$ but have to confirm that.

Produce a posterior for the survival curve:

```{r}
#| code-fold: true

d_post <- data.table(m1$draws(variables = c("surv0", "surv1"), format = "matrix"))
d_post <- melt(d_post, measure.vars = names(d_post))
d_post[variable %like% "surv0", trt := 0]
d_post[variable %like% "surv1", trt := 1]
d_fig <- copy(d_post)

d_fig[, x := gsub(".*\\[", "", variable)]
d_fig[, x := gsub("\\]", "", x)]
d_fig[, x := as.numeric(x)]

d_fig <- d_fig[
  , .(mu = mean(value),
      q_025 = quantile(value, prob = 0.025),
      q_975 = quantile(value, prob = 0.975)), keyby = .(trt, x)]
d_fig[, trt := factor(trt, levels = 0:1, labels = c("ctl", "trt"))]

ggplot(d_fig, aes(x = x, y = mu, group = trt, col = trt)) +
  geom_ribbon(aes(ymin = q_025, ymax = q_975, fill = trt), alpha = 0.1, col = NA) +
  geom_line() + 
  scale_y_continuous(limits = c(0.7, 1), breaks = seq(0.7, 1, by = 0.1)) +
  scale_color_discrete("") +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )
```

Estimate restricted mean survival time by group by integrating under the survival curve function for each draw from the posterior.

The RMST can be interpreted as the average survival time (i.e. time without the event, here being occurrence of RSV) during a defined time period ranging from time 0 to a specific follow-up time point.

```{r}
#| code-fold: true


# Define the function to integrate
integrand_1 <- function(
    x, mu, shape) {
  
  a = exp(mu)
  S = 1 / (1 + (x/a)^shape)
  S
  
}


d_post <- data.table(m1$draws(variables = c("gamma", "shape"), format = "matrix"))
names(d_post) <- c(paste0("gamma", 1:2), "shape")
i <- 1
m_rmst <- matrix(NA, ncol = 2, nrow = nrow(d_post))
for(i in 1:nrow(d_post)){
  m_rmst[i, 1] <- integrate(
    integrand_1, lower = 0, upper = 360,
    mu = d_post$gamma1[i], 
    shape = d_post$shape[i])$value  
  m_rmst[i, 2] <- integrate(
    integrand_1, lower = 0, upper = 360,
    mu = d_post$gamma1[i] + d_post$gamma2[i], 
    shape = d_post$shape[i])$value  
}



d_rmst <- data.table(m_rmst)
names(d_rmst) <- paste0(0:1)

rmst_diff <- d_rmst$`1` -  d_rmst$`0`


d_fig <- melt(d_rmst, measure.vars = names(d_rmst), variable.name = "trt")

d_fig[, trt := factor(trt, levels = 0:1, labels = c("ctl", "trt"))]

ggplot(d_fig, aes(x = value, group = trt, col = trt)) +
  geom_density() +
  scale_x_continuous("RMST") +
  scale_color_discrete("Treatment") +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )


names(d_rmst) <- paste0("rmst", 0:1)
d_rmst[, diff := rmst1 - rmst0]



```

From here, we could evaluate differences in the RMST between groups considering what level of improvement in the mean survival to 360 days would be warranted to adopt the treatment over the control.



<!-- ::: {#refs} -->
<!-- ::: -->
